---
title: "Getting Started with PricingBandits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PricingBandits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7.5, fig.height = 5)
```

`PricingBandits` implements the multi-armed bandit approaches to pricing
experiments from Weaver, Kumar, and Jain, *Nonparametric Pricing Bandits
Leveraging Informational Externalities to Learn the Demand Curve* (Marketing
Science). The setting is that a firm is trying to maximize profits while 
experimenting from a fixed set of candidate prices. Each consumer is presented
with a price and makes a decision whether to purchase if their WTP is greater
than that price. The algorithm observes only whether they purchase. The 
package's entry point is a single function, `PricingBandit()`.

```{r setup}
library(PricingBandits)
```

## The experiment environment

Everything about the demand environment is captured by one vector: the
consumers' **valuations** (willingness to pay), one draw per arriving consumer.
The package makes no assumption about where they come from — an analytic
distribution, an empirical CDF, transaction data, anything. A consumer buys
if and only if their valuation exceeds the posted price.

In this vignette, we use a right-skewed Beta(2, 9) population — a difficult case, 
because the revenue-maximizing price sits near the bottom of the price grid — with 
1,000 consumers. Every algorithm below is run against **this exact same sequence 
of consumers**, to minimize differences from luck of the draw.

```{r valuations, eval = FALSE}
set.seed(29)
valuations <- rbeta(1000, 2, 9)
prices     <- seq(10)/10          # ten candidate prices: 0.1, 0.2, ..., 1.0
```

## The `PricingBandit()` arguments, one by one

* **`valuations`** — numeric vector, one WTP draw per consumer. Its length sets
  the experiment length. Prices and valuations are assumed scaled to (0, 1]
  (rescale your data by its maximum price, and rescale back afterwards).
* **`prices`** — the candidate price grid (the bandit's arms), values in
  (0, 1]. Need not be evenly spaced.
* **`policy`** — which algorithm prices the consumers (see the table below).
* **`batch_size`** — how many consumers are served between policy updates
  (default 10). Larger batches mean fewer, cheaper updates but slower learning.
* **`num_knots`** — for the monotonic (`"-M"`) variants only: the number of
  knots of the basis expansion used to build monotone demand curves. **The
  default of 11 is deliberate and should be kept even for dense price grids**
  (e.g. 100 arms): with many more knots the truncated sampler operates in a
  near-degenerate, high-dimensional space and can break down.
* **`hetero`** — if `TRUE`, the observation-noise level at each price is
  re-estimated every update from the Gaussian-process posterior instead of
  being held at its conservative Bernoulli bound. Available for all GP
  variants.
* **`reset`** — if set to an integer `n`, the experiment history is wiped every
  `n` consumers. Useful when demand shifts over time (e.g. seasonality) and old
  observations mislead.
* **`timeout`** — seconds allowed for each truncated-sampling attempt in the
  monotonic fallback chain (default 5). Increase on slow machines to give the
  exact sampler more time; decrease to fail over to the cheaper approximations
  sooner.

The available policies:

| `policy` | Idea |
|----------|------|
| `"UCB"` | Upper Confidence Bound; every price learned independently |
| `"TS"` | Thompson Sampling with independent Beta posteriors per price |
| `"GP-UCB"`, `"GP-TS"` | Prices tied together through a Gaussian-process demand curve, so each observation informs *all* prices |
| `"GP-UCB-M"`, `"GP-TS-M"` | Additionally impose that demand is weakly decreasing in price — the sampled curves are monotone everywhere by construction |

## Running every algorithm on the same consumers

Each call returns a data frame with one row per consumer (`PricesTested`,
`PurchaseDecisions`), plus a `diagnostics` attribute with fallback counters.
We reset the seed before each run so the *policies'* randomness is reproducible
too, while the consumer sequence stays fixed.

```{r runs, eval = FALSE}
run <- function(policy, hetero = FALSE) {
  set.seed(1)
  PricingBandit(valuations, prices,
                policy     = policy,
                batch_size = 10,
                hetero     = hetero)
}

# Baselines: each arm learned independently
out_ucb  <- run("UCB")
out_ts   <- run("TS")

# Gaussian-process variants: the demand curve correlates the arms
out_gpucb <- run("GP-UCB")
out_gpts  <- run("GP-TS")

# Monotonic variants (basis-function construction; num_knots = 11 default)
out_gpucb_m <- run("GP-UCB-M")
out_gpts_m  <- run("GP-TS-M")

# Heterogeneous-noise versions of the monotonic algorithms
out_gpucb_m_h <- run("GP-UCB-M", hetero = TRUE)
out_gpts_m_h  <- run("GP-TS-M",  hetero = TRUE)
```

To judge performance we score each posted price by its *expected* revenue
`p * (1 - F(p))` under the true WTP distribution, and track cumulative revenue
as a percentage of what the true optimal price would have earned:

```{r metric, eval = FALSE}
expected_reward <- prices * (1 - pbeta(prices, 2, 9))
grid            <- seq(1e-6, 1, 1e-6)
true_optimal    <- max(grid * (1 - pbeta(grid, 2, 9)))

score <- function(out) {
  er <- expected_reward[match(out$PricesTested, prices)]
  cumsum(er) / (seq_along(er) * true_optimal) * 100
}
```

## Results

The results below were precomputed with exactly the code above (they ship with
the package so this vignette builds quickly).

```{r plot, echo = TRUE}
res <- readRDS("vignette_results.rds")$results

if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  res$family <- ifelse(grepl("TS", res$policy), "Thompson Sampling family",
                       "UCB family")
  res$variant <- ifelse(res$hetero, "heterogeneous noise", "standard")
  ggplot(res, aes(consumer, cum_pct_optimal, colour = policy,
                  linetype = variant)) +
    geom_line(linewidth = 0.6) +
    facet_wrap(~ family) +
    labs(x = "Consumers", y = "Cumulative revenue (% of true optimal)",
         colour = NULL, linetype = NULL,
         title = "All algorithms on the same 1,000 Beta(2,9) consumers") +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    theme(legend.position = "bottom")
} else {
  final <- res[res$consumer == 1000, c("label", "cum_pct_optimal")]
  final[order(-final$cum_pct_optimal), ]
}
```

The ordering reflects the paper's central result: exploiting the informational
externalities — first correlation across prices (GP), then monotonicity of
demand (the "-M" variants) — dramatically reduces the cost of learning,
especially in this hard case where the optimal price sits at the low end of
the grid.

Final standings after 1,000 consumers:

```{r table, echo = FALSE}
final <- res[res$consumer == 1000, c("label", "cum_pct_optimal")]
final <- final[order(-final$cum_pct_optimal), ]
knitr::kable(final, digits = 1, row.names = FALSE,
             col.names = c("Algorithm", "% of optimal (cumulative, 1000 consumers)"))
```

## Diagnostics

Each run counts how often its numerical fallback paths fired (hyperparameter
optimization failing back to priors, truncated-sampler timeouts, last-resort
samplers). In normal operation all counters are zero; a run with many
last-resort events is telling you the sampler struggled with your price grid.

```{r diag, eval = FALSE}
attr(out_gpts_m, "diagnostics")
```
